Tail associations in ecological variables and their impact on extinction risk

### 2020-03-23 MainText.Rmd
appendix fixed

The stochastic quantities $s_e(t)$ and $\tilde{f}_a(t)$ were generated 
to have the same overall association (Spearman's $\rho$ equal to $0.875$,
up to sampling variance), but different degrees and types of 
asymmetry of tail association in different simulations, as follows.
For a simulation of length $T$, we used methods that can 
generate $T$ independent pairs $(u(t),v(t))$ for $t=1,\ldots,T$ with
$\rho=0.875$ (up to sampling variance), but with either extreme left-,
moderate left-, moderate right-, or extreme right-tail association 
(e.g., Fig. \ref{fig_intro_cop}). The methods we used generated
$(u(t),v(t))$ such that $u(t)$ was uniformly distributed on $(0,1)$,
and likewise for $v(t)$. We then applied a monotonic transformation
to the $u(t)$ so that the resulting numbers were distributed according to
a beta distribution with parameters
$\alpha=3$ and $\beta=2$, 
and we used those values for the $s_e(t)$. We also applied a (different)
monotonic transformation to the $v(t)$ so that the resulting numbers were
distributed according to a gamma distribution with parameters $k=2$
and $\theta=1$, and we used those values for the $\tilde{f}_a(t)$. Because
the $u(t)$ and $v(t)$ were uniformly distributed, the inverse 
cumulative distribution functions of the desired beta and
gamma distributions were the transformations. Because these
were monotonic transformations, the Spearman correlation of 
$s_e(t)$ and $\tilde{f}_a(t)$ was the same as that of 
$u(t)$ and $v(t)$. The quantities $(u(t),v(t))$ for $t=1,\ldots,T$ were
generated as described in Appendix \ref{SM-makenoise}.

@@

<!--density-dependence metapopulation model-->
To investigate Q2 from the Introduction, we considered spatialized, stochastic versions
of six density dependent population models, the Ricker, Hassell, Maynard Smith, Pennycuick, Verhulst, 
and Austin-Brewer models. These models were all described by @cohen1995unexpected. 
The models are also listed in Table \ref{tab_model_summary}, and 
are analyzed in detail in Appendix \ref{SM-model_stability}, where 
original references for the models are also provided.
All models took the general form $\vec{P} (t+1) = D \lambda(t) \vec{P}(t)$,
where $\vec{P}(t)$ is a vector of population densities, 
the $i^{\text{th}}$ component $P_i(t)$ representing
the population density in the $i^{\text{th}}$ habitat patch ($i=1,2,\ldots,n$) at time $t$. 
The $n \times n$ matrix $\lambda(t)$ was diagonal with $i^{\text{th}}$ diagonal entry 
$\lambda_i(t)$ depending on the model used (Table \ref{tab_model_summary}). The matrix $D$
was an $n \times n$ dispersal matrix. For simplicity, 
we assumed the $n$ habitat patches were arranged, evenly spaced, in a line.
We considered both "local" and "global" dispersal among patches in different simulations. 
For local dispersal, a fraction $d$ of 
each population dispersed during each time step, equally distributed to
the two or one nearest neighbors. For global dispersal the same fraction $d$ dispersed
equally to the other $n-1$ patches. When $d$ was $0$, dispersal did not occur and 
rescue effects were not possible.

We obtained our list of classic population models (which we then spatialized and made 
stochastic, as above) from @cohen1995unexpected. 
Model parameters were selected so that varying the growth parameter, $r$, across a 
range, while keeping the other parameters fixed at selected values, 
caused the  deterministic, one-patch versions of the models to transition from dynamics
showing an undercompensatory (i.e., monotonic) approach to an equilibrium; to
dynamics showing an overcompensatory (i.e., oscillatory) approach to an equilibrium. 
The transition value of $r$ was denoted $r_c$. Appendix \ref{SM-model_stability}
summarizes stability analyses of the models.
Whereas @cohen1995unexpected
was interested in studying chaotic model behavior and therefore chose parameters
for which the models exhibited chaotic dynamics, we were interested in non-chaotic 
behaviors, which seem likely to be more common in real populations. 
Two models that @cohen1995unexpected 
used were excluded from our analyses because their functional forms complicated 
the analyses we wished to perform. First, the equilibrium of the 
Malthus-Condorcet-Mill model [@cohen1995unexpected] could not be determined 
analytically, making downstream analyses difficult for that model. Second, for
the Varley model and for the portions of parameter space we explored, 
varying $r$ produced no transition from under- to overcompensatory dynamics.

We simulated our metapopulation models using environmental noises $\varepsilon_i(t)$ (Table \ref{tab_model_summary}), 
independent through time, that were generated to have the same overall associations 
between all patches and in all simulations (Spearman's $\rho$ again `r srho_common`, up to sampling variance), 
but with different degrees and types of asymmetry of tail associations 
between patches in different simulations, as follows. For a simulation of length $T$,
we used methods that can generate $T$ independent $n$-tuples $(u_1(t),\ldots,u_n(t))$
for $t=1,\ldots,T$ with $\rho=$ `r srho_common` between $u_i(t)$ and $u_j(t)$ for all $i \neq j$
(up to sampling variance), but with either extreme left-, moderate left-, moderate right-,
or extreme right-tail association between all pairs of locations (as in Fig. \ref{fig_intro_cop}). 
Pairs $u_i(t)$ and $u_j(t)$ looked like Fig. \ref{fig_intro_cop} and were structured, statistically,
in the same way as the stochastic quantities $u(t)$ and $v(t)$ used for the non-spatial 
simulations described above. For any $i$, $u_i(t)$ was again uniformly distributed on 
the interval $(0,1)$. To obtain the values $\varepsilon_i(t)$, the $u_i(t)$ were 
transformed so that the resulting numbers were
normally distributed with mean $0$ and standard deviation $\sigma$ (Table \ref{tab_model_summary}), 
for all $i$. This was 
accomplished with the inverse cumulative distribution function of the normal distribution
with mean $0$ and standard deviation $\sigma$. Because this was a monotonic transformation,
the Spearman correlation between $\varepsilon_i(t)$ and $\varepsilon_j(t)$ was the same 
as that between $u_i(t)$ and $u_j(t)$; likewise partial Spearman correlations were the same. 
The quantities $(u_1(t),\ldots,u_n(t))$
for $t=1,\ldots,T$ were generated as described in Appendix \ref{SM-makenoise}.

@@

<!--test hypothesis from Q2 : ext risk with density-dependence metapopulation model with r from different regime-->
To answer (Q2) of the Introduction, 
we estimated extinction risk for each metapopulation model, using each form of tail 
association for the environmental noise,
using values of $r$ that included under- and overcompensatory dynamics,
as indicated above. We chose values of $r$ within a range, the
range specified as follows. The second part of Q2 (about how density dependence mediates the
influence of asymmetric tail associations on extinction risk) was addressed by comparing
results for under- and overcompensatory model regimes.
Defining $r_\text{min}$ as the minimum value of $r$ 
for which $P_e$ was non-negative; and defining
$r_\text{bf}$ as the value of $r$ at which a bifurcation occurs, the range of
$r$ used spanned 95\% of the interval ($r_\text{min},r_\text{bf}$). For details, see Appendix 
\ref{SM-model_stability}. The range of $r$ encompassed $r_c$, so included 
a transition from under- to overcompensatory dynamics. For each value of $r$
used within the range, we calculated extinction risk as indicated above.
Dispersal parameter values $d= 0,0.1,0.2,\ldots,0.5$ were also used in different 
simulations. We plotted extinction risk as a function of $d$ for several
fixed values of $r$, some producing under- and some overcompensatory dynamics.
For $d=0$ we also plotted extinction risk as a function of $r$. All code for 
this project is available at https://github.com/sghosh89/ERC.

@@

<!--Summary tables for 6 nonlinear models used-->
```{r read_tab_model_summary,echo=F, results='asis',message=F}
library(kableExtra)
library(dplyr)
dt<-readRDS("./Results/tab_model_summary.RDS")
knitr::kable(dt, escape=F,
             format="latex",digits=3, align="c",table.env='table*',
             caption = "Summary of the six density-dependent stochastic metapopulation models we used. The population growth rate in habitat patch $i$ at time $t$, $\\lambda_i(t)$ (second column), is $P_i(t+1)/P_i(t)$, and defines how the model works within each patch. Here $P_i(t)$ is the population in patch $i$ at time $t$. The growth rate depends on model parameters, $P_i(t)$, and the environment $\\varepsilon_i(t)$ in patch $i$ at time $t$. $\\varepsilon_i(t)$ was normally distributed with mean $0$ and standard deviation $\\sigma$, and values were generated independently across times. $P_e$ is the equilibrium of the deterministic, one-patch version of the model. $r_\\text{min}$ is the smallest $r$, given fixed values of the other parameters, for which $P_e$ was non-negative. $r_c$ is the value of $r$ for which a transition occurs from under- to overcompensatory dynamics (see main text). $r_\\text{bf}$ is the value of $r$ for which a transition occurs from a stable to an unstable equilibrium (a bifurcation; see main text). For extinction-risk simulations, we used $r$ such that $r_\\text{min}<r<r_\\text{bf}$. Dispersal between patches was also implemented (see main text). Formulas are given for $P_e$ but specific values which pertain for the listed parameters are given for $r_\\text{min}$, $r_c$ and $r_\\text{bf}$. See Appendix \\ref{SM-model_stability} for the formulas that gave these values.\\label{tab_model_summary}",
             booktabs=T, linesep = "\\addlinespace"
             )%>%column_spec(c(1,3,5,6,7),width="1.4 cm")%>%kable_styling(latex_options = "basic",position = "center")%>%column_spec(2, width = "3.8 cm")
```

>>>>

The stochastic quantities $s_e(t)$ and $\tilde{f}_a(t)$ were generated 
to have the same overall association (Spearman's $\rho$ equal to $0.875$,
up to sampling variance), but different degrees and types of 
asymmetry of tail association in different simulations, as follows.
For a simulation of length $T$, we used methods that can 
generate $T$ independent pairs $(u(t),v(t))$ for $t=1,\ldots,T$ with
$\rho=0.875$ (up to sampling variance), but with either extreme left-,
moderate left-, moderate right-, or extreme right-tail association 
(e.g., Fig. \ref{fig_intro_cop}). The methods we used generated
$(u(t),v(t))$ such that $u(t)$ was uniformly distributed on $(0,1)$,
and likewise for $v(t)$. We then applied a monotonic transformation
to the $u(t)$ so that the resulting numbers were distributed according to
a beta distribution with parameters
$\alpha=3$ and $\beta=2$, 
and we used those values for the $s_e(t)$. We also applied a (different)
monotonic transformation to the $v(t)$ so that the resulting numbers were
distributed according to a gamma distribution with parameters $k=2$
and $\theta=1$, and we used those values for the $\tilde{f}_a(t)$. Because
the $u(t)$ and $v(t)$ were uniformly distributed, the inverse 
cumulative distribution functions of the desired beta and
gamma distributions were the transformations. Because these
were monotonic transformations, the Spearman correlation of 
$s_e(t)$ and $\tilde{f}_a(t)$ was the same as that of 
$u(t)$ and $v(t)$. The quantities $(u(t),v(t))$ for $t=1,\ldots,T$ were
generated as described in Appendix, section \ref{SM-makenoise}.

@@

<!--density-dependence metapopulation model-->
To investigate Q2 from the Introduction, we considered spatialized, stochastic versions
of six density dependent population models, the Ricker, Hassell, Maynard Smith, Pennycuick, Verhulst, 
and Austin-Brewer models. These models were all described by @cohen1995unexpected. 
The models are also listed in Table \ref{tab_model_summary}, and 
are analyzed in detail in Appendix, section \ref{SM-model_stability}, where 
original references for the models are also provided.
All models took the general form $\vec{P} (t+1) = D \lambda(t) \vec{P}(t)$,
where $\vec{P}(t)$ is a vector of population densities, 
the $i^{\text{th}}$ component $P_i(t)$ representing
the population density in the $i^{\text{th}}$ habitat patch ($i=1,2,\ldots,n$) at time $t$. 
The $n \times n$ matrix $\lambda(t)$ was diagonal with $i^{\text{th}}$ diagonal entry 
$\lambda_i(t)$ depending on the model used (Table \ref{tab_model_summary}). The matrix $D$
was an $n \times n$ dispersal matrix. For simplicity, 
we assumed the $n$ habitat patches were arranged, evenly spaced, in a line.
We considered both "local" and "global" dispersal among patches in different simulations. 
For local dispersal, a fraction $d$ of 
each population dispersed during each time step, equally distributed to
the two or one nearest neighbors. For global dispersal the same fraction $d$ dispersed
equally to the other $n-1$ patches. When $d$ was $0$, dispersal did not occur and 
rescue effects were not possible.

We obtained our list of classic population models (which we then spatialized and made 
stochastic, as above) from @cohen1995unexpected. 
Model parameters were selected so that varying the growth parameter, $r$, across a 
range, while keeping the other parameters fixed at selected values, 
caused the  deterministic, one-patch versions of the models to transition from dynamics
showing an undercompensatory (i.e., monotonic) approach to an equilibrium; to
dynamics showing an overcompensatory (i.e., oscillatory) approach to an equilibrium. 
The transition value of $r$ was denoted $r_c$. Appendix, section \ref{SM-model_stability}
summarizes stability analyses of the models.
Whereas @cohen1995unexpected
was interested in studying chaotic model behavior and therefore chose parameters
for which the models exhibited chaotic dynamics, we were interested in non-chaotic 
behaviors, which seem likely to be more common in real populations. 
Two models that @cohen1995unexpected 
used were excluded from our analyses because their functional forms complicated 
the analyses we wished to perform. First, the equilibrium of the 
Malthus-Condorcet-Mill model [@cohen1995unexpected] could not be determined 
analytically, making downstream analyses difficult for that model. Second, for
the Varley model and for the portions of parameter space we explored, 
varying $r$ produced no transition from under- to overcompensatory dynamics.

We simulated our metapopulation models using environmental noises $\varepsilon_i(t)$ (Table \ref{tab_model_summary}), 
independent through time, that were generated to have the same overall associations 
between all patches and in all simulations (Spearman's $\rho$ again `r srho_common`, up to sampling variance), 
but with different degrees and types of asymmetry of tail associations 
between patches in different simulations, as follows. For a simulation of length $T$,
we used methods that can generate $T$ independent $n$-tuples $(u_1(t),\ldots,u_n(t))$
for $t=1,\ldots,T$ with $\rho=$ `r srho_common` between $u_i(t)$ and $u_j(t)$ for all $i \neq j$
(up to sampling variance), but with either extreme left-, moderate left-, moderate right-,
or extreme right-tail association between all pairs of locations (as in Fig. \ref{fig_intro_cop}). 
Pairs $u_i(t)$ and $u_j(t)$ looked like Fig. \ref{fig_intro_cop} and were structured, statistically,
in the same way as the stochastic quantities $u(t)$ and $v(t)$ used for the non-spatial 
simulations described above. For any $i$, $u_i(t)$ was again uniformly distributed on 
the interval $(0,1)$. To obtain the values $\varepsilon_i(t)$, the $u_i(t)$ were 
transformed so that the resulting numbers were
normally distributed with mean $0$ and standard deviation $\sigma$ (Table \ref{tab_model_summary}), 
for all $i$. This was 
accomplished with the inverse cumulative distribution function of the normal distribution
with mean $0$ and standard deviation $\sigma$. Because this was a monotonic transformation,
the Spearman correlation between $\varepsilon_i(t)$ and $\varepsilon_j(t)$ was the same 
as that between $u_i(t)$ and $u_j(t)$; likewise partial Spearman correlations were the same. 
The quantities $(u_1(t),\ldots,u_n(t))$
for $t=1,\ldots,T$ were generated as described in Appendix, section \ref{SM-makenoise}.

@@

<!--test hypothesis from Q2 : ext risk with density-dependence metapopulation model with r from different regime-->
To answer (Q2) of the Introduction, 
we estimated extinction risk for each metapopulation model, using each form of tail 
association for the environmental noise,
using values of $r$ that included under- and overcompensatory dynamics,
as indicated above. We chose values of $r$ within a range, the
range specified as follows. The second part of Q2 (about how density dependence mediates the
influence of asymmetric tail associations on extinction risk) was addressed by comparing
results for under- and overcompensatory model regimes.
Defining $r_\text{min}$ as the minimum value of $r$ 
for which $P_e$ was non-negative; and defining
$r_\text{bf}$ as the value of $r$ at which a bifurcation occurs, the range of
$r$ used spanned 95\% of the interval ($r_\text{min},r_\text{bf}$). For details, see Appendix, section 
\ref{SM-model_stability}. The range of $r$ encompassed $r_c$, so included 
a transition from under- to overcompensatory dynamics. For each value of $r$
used within the range, we calculated extinction risk as indicated above.
Dispersal parameter values $d= 0,0.1,0.2,\ldots,0.5$ were also used in different 
simulations. We plotted extinction risk as a function of $d$ for several
fixed values of $r$, some producing under- and some overcompensatory dynamics.
For $d=0$ we also plotted extinction risk as a function of $r$. All code for 
this project is available at https://github.com/sghosh89/ERC.

@@

<!--Summary tables for 6 nonlinear models used-->
```{r read_tab_model_summary,echo=F, results='asis',message=F}
library(kableExtra)
library(dplyr)
dt<-readRDS("./Results/tab_model_summary.RDS")
knitr::kable(dt, escape=F,
             format="latex",digits=3, align="c",table.env='table*',
             caption = "Summary of the six density-dependent stochastic metapopulation models we used. The population growth rate in habitat patch $i$ at time $t$, $\\lambda_i(t)$ (second column), is $P_i(t+1)/P_i(t)$, and defines how the model works within each patch. Here $P_i(t)$ is the population in patch $i$ at time $t$. The growth rate depends on model parameters, $P_i(t)$, and the environment $\\varepsilon_i(t)$ in patch $i$ at time $t$. $\\varepsilon_i(t)$ was normally distributed with mean $0$ and standard deviation $\\sigma$, and values were generated independently across times. $P_e$ is the equilibrium of the deterministic, one-patch version of the model. $r_\\text{min}$ is the smallest $r$, given fixed values of the other parameters, for which $P_e$ was non-negative. $r_c$ is the value of $r$ for which a transition occurs from under- to overcompensatory dynamics (see main text). $r_\\text{bf}$ is the value of $r$ for which a transition occurs from a stable to an unstable equilibrium (a bifurcation; see main text). For extinction-risk simulations, we used $r$ such that $r_\\text{min}<r<r_\\text{bf}$. Dispersal between patches was also implemented (see main text). Formulas are given for $P_e$ but specific values which pertain for the listed parameters are given for $r_\\text{min}$, $r_c$ and $r_\\text{bf}$. See Appendix, section \\ref{SM-model_stability} for the formulas that gave these values.\\label{tab_model_summary}",
             booktabs=T, linesep = "\\addlinespace"
             )%>%column_spec(c(1,3,5,6,7),width="1.4 cm")%>%kable_styling(latex_options = "basic",position = "center")%>%column_spec(2, width = "3.8 cm")
```
